---
title: Get started with limpca.
description: >
  Get started with limpca.
output: rmarkdown::html_vignette
bibliography: references.bib  
vignette: >
  %\VignetteIndexEntry{Get started with limpca.}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    out.width = "70%",
    fig.align = "center",
    eval = TRUE
)
rm(list = ls())
library("limpca")

if (!requireNamespace("SummarizedExperiment", quietly = TRUE)) {
    stop("Install 'SummarizedExperiment' to knit this vignette")
}
library(SummarizedExperiment)
```  


<!-- badges: start -->
[![R-CMD-check](https://github.com/ManonMartin/limpca/actions/workflows/check-standard.yaml/badge.svg)](https://github.com/ManonMartin/limpca/actions/workflows/check-standard.yaml)
<!-- badges: end -->

# Installation

```{r, eval=FALSE}
remotes::install_github("ManonMartin/limpca", dependencies = TRUE)
library("limpca")
```

Note that if you would like to build the vignettes, you have to install `BiocStyle` (from [Bioconductor](https://www.bioconductor.org/)) and `rmarkdown` packages before installing `limpca` with the following command: `remotes::install_github("ManonMartin/limpca", dependencies = TRUE, build_vignettes = TRUE)`.

For any enquiry, you can send an email to the package authors: [bernadette.govaerts@uclouvain.be](mailto:bernadette.govaerts@uclouvain.be) ; [michel.thiel@uclouvain.be](mailto:michel.thiel@uclouvain.be) or [manon.martin@uclouvain.be](mailto:manon.martin@uclouvain.be)

# Short application on the `UCH` dataset

## Data object

In order to use the limpca core functions, the data need to be formatted as a list (informally called an lmpDataList) with the following elements: `outcomes` (multivariate matrix), `design` (data.frame) and `formula` (character string). 
The `UCH` data set is already formatted appropriately and can be loaded from `limpca` with the `data` function.

```{r}
data("UCH")
str(UCH)
```



Alternatively, the lmpDataList can be created with the function `data2LmpDataList` :

- from scratch:

```{r, message = TRUE}
UCH2 <- data2LmpDataList(
   outcomes = UCH$outcomes,
   design = UCH$design, 
   formula = UCH$formula
 )
```

- or from a `SummarizedExperiment`:

```{r, message = TRUE}
se <- SummarizedExperiment(
   assays = list(
     counts = t(UCH$outcomes)), colData = UCH$design,
   metadata = list(formula = UCH$formula)
 )

UCH3 <- data2LmpDataList(se, assay_name = "counts")
```

`SummarizedExperiment` is a generic data container that stores rectangular matrices of experimental results. See @SummarizedExperiment23 for more information.


## Data visualisation

The design can be visualised with `plotDesign()`.

```{r dataVisu}
# design
plotDesign(
    design = UCH$design, x = "Hippurate",
    y = "Citrate", rows = "Time",
    title = "Design of the UCH dataset"
)

# row 3 of outcomes
plotLine(
    Y = UCH$outcomes,
    title = "H-NMR spectrum",
    rows = c(3),
    xlab = "ppm",
    ylab = "Intensity"
)
```

## PCA

```{r PCA}
ResPCA <- pcaBySvd(UCH$outcomes)
pcaScreePlot(ResPCA, nPC = 6)
pcaScorePlot(
    resPcaBySvd = ResPCA, axes = c(1, 2),
    title = "PCA scores plot: PC1 and PC2",
    design = UCH$design,
    color = "Hippurate", shape = "Citrate",
    points_labs_rn = FALSE
)
```

## Model estimation and effect matrix decomposition

```{r modelEst}
# Model matrix generation
resMM <- lmpModelMatrix(UCH)

# Model estimation and effect matrices decomposition
resEM <- lmpEffectMatrices(resMM)
```

## Effect matrix test of significance and importance measure

```{r effectImpSign}
# Effects importance
resEM$varPercentagesPlot

# Bootstrap tests
resBT <- lmpBootstrapTests(resLmpEffectMatrices = resEM, nboot = 100)
resBT$resultsTable
```


## ASCA decomposition

```{r ASCA}
# ASCA decomposition
resASCA <- lmpPcaEffects(resLmpEffectMatrices = resEM, method = "ASCA")

# Scores Plot for the hippurate
lmpScorePlot(resASCA,
    effectNames = "Hippurate",
    color = "Hippurate", shape = "Hippurate"
)

# Loadings Plot for the hippurate
lmpLoading1dPlot(resASCA,
    effectNames = c("Hippurate"),
    axes = 1, xlab = "ppm"
)

# Scores ScatterPlot matrix
lmpScoreScatterPlotM(resASCA,
    PCdim = c(1, 1, 1, 1, 1, 1, 1, 2),
    modelAbbrev = TRUE,
    varname.colorup = "Citrate",
    varname.colordown = "Time",
    varname.pchup = "Hippurate",
    varname.pchdown = "Time",
    title = "ASCA scores scatterplot matrix"
)
```


# sessionInfo
```{r}
sessionInfo()
```

