---
title: "three-way ANOVA with fixed effects: UCH data"
author: "Benaiche Nadia, Sébastien Franceschini, Martin Manon, Thiel Michel, Govaerts Bernadette"
date: '`r format(Sys.time(), "%B %d, %Y")`'
package: limpca
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 4
    toc_float: true
vignette: >
   %\VignetteIndexEntry{Analysis of the UCH dataset with limpca}
   %\VignetteEncoding{UTF-8}
   %\VignetteEngine{knitr::rmarkdown}
editor_options: 
   chunk_output_type: console
references:
- id: Thiel2017
  title: ASCA+ and APCA+ Extensions of ASCA and APCA in the analysis of unbalanced multifactorial designs
  author:
  - family: Thiel
    given: Michel
  - family: Feraud
    given: Baptiste
  - family: Govaerts
    given: Bernadette
  container-title: Journal of Chemometrics
  type: article-journal
  issued:
    year: 2017
- id: Guisset2019
  title: Comparison of PARAFASCA, AComDim, and AMOPLS approaches in the multivariate GLM modelling of multi-factorial designs
  author:
  - family: Guisset
    given: Severine
  - family: Martin
    given: Manon
  - family: Govaerts
    given: Bernadette
  container-title: Chemometrics and Intelligent Laboratory Systems
  type: article-journal
  issued:
    year: 2019
---

```{r Setup,include=FALSE}
# require(knitr)
knitr::opts_chunk$set(message=FALSE,warning=TRUE,comment=NA, crop = NULL,
                      tidy.opts = list(width.cutoff = 60), tidy = TRUE)
rm(list=ls())
```

```{r}
library(pander)
library(ggplot2)
library(gridExtra)
```


# Introduction

The purpose of this vignette is to show the possibilities offered by the `limpca` package. 

limpca stands for linear modeling of high-dimensional designed data based on ASCA/APCA family of methods. This package was created to analyse models with high-dimensional data and a multi-factor design of experiment and provide an implementation of ASCA and APCA derived methods.  

The model used in this example is a three-way ANOVA with fixed effects. This document presents all the usual steps of the analysis, from importing the data to visualising the results. Details on the methods used can be found in the articles of @Thiel2017 and @Guisset2019.

# Installation

The package is actually in its development stage and is available on GitHub: <https://github.com/ManonMartin/limpca>. It can be installed via the `remotes::install_github()` function. The package needs to be loaded afterwards.

```{r Install, eval=FALSE}
if (!requireNamespace("remotes"))
install.packages(pkgs="remotes", dependencies="Depends")
remotes::install_github("ManonMartin/limpca", 
                        dependencies = TRUE)
```

Note that if you would like to build the vignettes, you have to install `BiocStyle` (from [Bioconductor](https://www.bioconductor.org/)) and `rmarkdown` packages before installing `limpca` with the following command: `remotes::install_github("ManonMartin/limpca", dependencies = TRUE, build_vignettes = TRUE)`.


```{r Load, results=FALSE, message=FALSE}
library("limpca")
```

# Data importation

Before any analysis, the `UCH` data set needs to be loaded. The limpca package contains the data set and it can be loaded with the `data()` function. The `load()` function is also useful to import your own data.

```{r Data Importation}
data("UCH")
UCH$formula <- "outcomes ~ Hippurate + Citrate + Time + Hippurate:Citrate + Time:Hippurate + Time:Citrate + Hippurate:Citrate:Time"
```

# Data exploration

The UCH (Urine-Citrate-Hippurate) data set is described in @Thiel2017 and @Guisset2019 and is issued form a metabolomics experiment.  In this experiment, 36 samples of a pool of rat urine samples were spiked with two molecules Citrate and Hippurate according to a $3^2$ full factorial design in the quantities of these two molecules.  The spiked samples were analyzed by ^1^H NMR at two different time after defrozing and over two days.  Two of the spectra where finally missing at the end of the experiment.    

The UCH data set is a list containing 3 elements: 

* an `outcomes` matrix with 34 observations of 600 response variables representing the spectra from the \textsuperscript^1H-NMR$ spectroscopy, 
* a `design` matrix with 34 observations and 4 explanatory variables and 
* a `formula` for the General Linear Model (GLM) used. 

For the purpose of this example, only 3 factors of interest will be studied : Quantities of Hippurate and Citrate and Time after defrozing.

## Design

The design matrix contains the information about each observation for the four variables: Hippurate, Citrate, Day and Time. Only 3 of these variables are used in the model. The function `plotDesign` is useful to observe the design.

```{r Design visualization}

pander(head(UCH$design))
plotDesign(design = UCH$design, x = "Hippurate", 
           y = "Citrate", rows = "Time",
           title = "Design of the UCH dataset")

```

This plot confirms that the design is a full 3x3x2 factorial design replicated twice with 2 missing values.  The design is then not balanced. 

## Outcomes visualization

The 600 response (`outcomes`) variables represent, for each observation, the intensities of the ^1^H NMR spectra.   These spectra can be visualized by the `plotLine` function.

### `plotLine` function

Here, annotation are added to the `ggplot` in order to highlight the Hippurate (green) and Citrate (red) peaks. 

```{r Spectrum visualization}
cit_peaks <- annotate("rect", xmin=c(2.509), xmax=c(2.709), 
              ymin=-Inf, ymax=Inf, alpha=0.2, 
              fill=c("tomato")) 

hip_peaks <- annotate("rect", xmin=c(7.458,3.881), xmax=c(7.935,4.041), 
              ymin=-Inf, ymax=Inf, alpha=0.2, 
              fill=c("yellowgreen"))

p1 <- plotLine(Y = UCH$outcomes,
         title = "H-NMR spectrum",
         rows = c(3),
         xlab = "ppm",
         ylab = "Intensity")

p1 + cit_peaks + hip_peaks
  
```

### `plotScatter` function

`plotScatter` function allows to visualize the values of two outcomes variables in highlighting with colors or markers the values of the design factors.  Here, it is used to show that the $3^2$ factorial design can be recovered from the intensities of the Hippurate and Citrate peaks in the spectra.  

```{r}

# xy corresponds to citrate (453) and hippurate peaks (369)
plotScatter(Y = UCH$outcomes,
            xy = c(453, 369),
            design = UCH$design,
            color = "Hippurate", 
            shape = "Citrate")
# Or
plotScatter(Y = UCH$outcomes,
            xy = c("2.6092056","3.9811536"),
            design = UCH$design,
            color = "Hippurate", 
            shape = "Citrate")

```

### `plotScatterM` function

`plotScatter` function allows to visualize the values of a series of outcomes variables in highlighting with colors or markers the values of the design factors.  It is done here for all peaks of Citrate et Hippurate

```{r}

plotScatterM(Y = UCH$outcomes, cols = c(133, 145, 150, 369, 453), 
             design = UCH$design,varname.colorup = "Hippurate", 
             varname.colordown = "Citrate")

```

### `plotMeans` function

`plotMeans` allows finally to visualize the mean values of a response variable for different levels of the design factors.  Here we show the evolution of the Citrate peak higth with respect to the three design factors of interest. 
Note that the results of this function must be interpreted with caution when designs are unbalanced.

```{r}

plotMeans(Y = UCH$outcomes,
          design = UCH$design,
          cols = c(453),
           x = c("Citrate"),
          w = c("Hippurate"),
          z = c("Time"),
          ylab = "Intensity",
          title=c("Mean reponse for main Citrate peak"))

```

## PCA {#pca}

The function `pcaBySvd` is useful to compute a Principal Component Analysis (PCA) decomposition of the `outcomes` matrix. The linked usefull functions are :

* `pcaScreePlot` to obtaine a scree plot
* `pcaLoading1dPlot` for the loading plots
* `pcaScorePlot` for the score plots

```{r PCA}

ResPCA = pcaBySvd(UCH$outcomes)
pcaScreePlot(ResPCA, nPC = 6) 

```

The score plots show a quite effect of the three design factors which will be more clearly highlighted by ASCA and APCA.  

```{r Scores}

pcaScorePlot(resPcaBySvd = ResPCA, axes = c(1,2), 
             title = "PCA scores plot: PC1 and PC2", 
             design = UCH$design,
             color = "Hippurate", shape = "Citrate",
             points_labs_rn = FALSE)

pcaScorePlot(resPcaBySvd = ResPCA, axes = c(1,2), 
             title = "PCA scores plot: PC1 and PC2", 
             design = UCH$design,
             color = "Time", shape = "Hippurate",
             points_labs_rn = FALSE)

pcaScorePlot(resPcaBySvd = ResPCA, axes = c(3,4), 
             title = "PCA scores plot: PC3 and PC4", 
             design = UCH$design,
             color = "Time", shape = "Citrate",
             points_labs_rn = FALSE)

```

In the first two loading plots a mix of Citrate and Hippurate peaks already appear.  

```{r Loadings}

p2 <- pcaLoading1dPlot(resPcaBySvd = ResPCA, axes = c(1,2), 
               title = "PCA loadings plot UCH", xlab = "ppm", 
               ylab = "Intensity")

p2 +  hip_peaks + cit_peaks

```

# Model estimation and effect matrix decomposition

## Model formula

The formula of the ANOVA-GLM model used in this analysis is the 3 ways crossed ANOVA model:

```{r Formula}
UCH$formula
```

## Model matrix generation

The first step of ASCA+ is to build the (GLM) model matrix from the experimental design matrix and the model. Each factor is reencoded with multiple binary variables using  `contr.sum` coding. The model matrix is a \emph(34xp) with p being the total number of parameter in the ANOVA model for one response.

The function `lmpModelMatrix()` encodes the design matrix as a model matrix.

```{r ModelMatrix}

resLmpModelMatrix = lmpModelMatrix(UCH)
pander::pander(head(resLmpModelMatrix$modelMatrix))
```

## Model estimation and effect matrices decomposition

`lmpEffectMatrices()` is the used to estimate the GLM model and decompose the outcomes matrix effect matrices for every model term.  This function calculates also type III effect contributions (in %) and generates a linked `barpot`.  

```{r EffectMatrices}
resLmpEffectMatrices = lmpEffectMatrices(resLmpModelMatrix)
```

## Effects importance

The contributions from each effect is outputted from `lmpEffectMatrices`.

```{r}
pander(resLmpEffectMatrices$variationPercentages)
resLmpEffectMatrices$varPercentagesPlot
```

## Bootstrap tests and quantification of effects importance

`lmpBootstrapTests()` allows to apply parametric bootstrap test to determine whether an effect is significant or not.  Use first a small value of `nboot` (e.g. nboot=100) to develop your code and increase then it (e.g. nboot=1000) in ordre to get an accurate value for the p-values.    

```{r Bootstrap}

resLmpBootstrapTests = 
  lmpBootstrapTests(resLmpEffectMatrices = resLmpEffectMatrices, 
                                         nboot=100)

# Print P-values
pander::pander(t(resLmpBootstrapTests$resultsTable))

```



# ASCA/APCA/ASCA-E decomposition

ASCA/APCA/ASCA-E decomposition allow to represent the information from the effect matrices in a space of reduced dimensions through PCA. The function `lmpPcaEffects()` has a method argument to define which method to use, namely `ASCA`, `APCA` or `ASCA-E`.

## ASCA

The ASCA method performs PCA on the pure effect matrices.

```{r ASCA PCA}
resASCA = lmpPcaEffects(resLmpEffectMatrices = resLmpEffectMatrices, 
                        method="ASCA", 
                        combineEffects = list(c("Hippurate", "Time", 
                                                "Hippurate:Time")))
```

### Contributions

The contribution of each principal component of the effects is estimated and reported in different tables and plots with the function `lmpContributions()`. 

```{r ASCA Contrib}
resLmpContributions = lmpContributions(resASCA)
```

The tables are:

- `totalContribTable`: Table of the **contribution of each effect to the total variance in percentage** as outputted from `lmpEffectMatrices`.
```{r}
pander::pander(resLmpContributions$totalContribTable)
```

- `effectTable`: Table of the **percentage of variance explained by each Principal Component in each model effect decomposition**.

```{r}
pander::pander(resLmpContributions$effectTable)
```

- `combinedEffectTable`: Equivalent of the previous `effectTable` but for the *combination of effects* mentioned in `lmpPcaEffects()`, here for `Hippurate+Time+Hippurate:Time`.
```{r}
pander::pander(resLmpContributions$combinedEffectTable)
```

- `contribTable`: Table of the **percentage of variance explained by each Principal Component of each effect reported to the percentage contribution of the given effect to the total variance**. This table gives an overview of the importance of each PC regardless of the effects.


```{r}
pander::pander(resLmpContributions$contribTable)
```

- Moreover `lmpContributions()` also produces a barplot either with the ordered contributions per effect (`plotTotal`) or across all the PCs of the different effects (`plotContrib`).


```{r}
pander("Ordered contributions per effect:")
resLmpContributions$plotTotal

pander("For each PC of the different effects:")
resLmpContributions$plotContrib
```

In the following analysis, we will focus only on the first PC of the three main effects, the interaction `Hippurate:Time` and the residuals since the other PCs and effects account for less than 1% of the total variance.


### Scores and loadings Plots

The loadings can be represented either on a line plot with the function `lmpLoading1dPlot()` to conveniently compare them with the original spectral profiles or on a scatterplot with the function `lmpLoading2dPlot()`. 



Here we create an object including the loading plots for all the effects included in the model as well as the combined effect.

```{r, include=TRUE}
all_loadings_pl <- lmpLoading1dPlot(resASCA,
                              effectNames = c("Hippurate", "Citrate","Time",
                                              "Hippurate:Time",
                                              "Hippurate+Time+Hippurate:Time",
                                              "Residuals"),
                              axes = 1, xlab = "ppm")
```

The score matrices are represented two components at a time on a scatterplot with the function `lmpScorePlot()`.   

#### Main effects

The scores and loadings of the main effects included in the model are represented below.

```{r ASCA ScoresXY}
# Hippurate
hip_scores_pl <- lmpScorePlot(resASCA, effectNames = "Hippurate", 
             color = "Hippurate", shape = "Hippurate")

hip_loadings_pl <- all_loadings_pl$Hippurate +  hip_peaks

grid.arrange(hip_scores_pl,hip_loadings_pl, ncol=2)
  
# Citrate
cit_scores_pl <- lmpScorePlot(resASCA, effectNames = "Citrate", 
             color = "Citrate", shape = "Citrate")
cit_loadings_pl <- all_loadings_pl$Citrate +  cit_peaks

grid.arrange(cit_scores_pl,cit_loadings_pl, ncol=2)

# Time
tim_scores_pl <- lmpScorePlot(resASCA, effectNames = "Time", color = "Time", 
                              shape = "Time")

time_peaks <- annotate("rect", xmin=c(5.955364), xmax=c(6.155364), 
              ymin=-Inf, ymax=Inf, alpha=0.2, 
              fill=c("royalblue")) 

tim_loadings_pl <- all_loadings_pl$Time +  time_peaks

grid.arrange(tim_scores_pl,tim_loadings_pl, ncol=2)

```

Hippurate and Citrate peaks are clearly and separately retrieved in the respective loading plots compares to the classical PCA (see Section \@ref(pca)) where peaks of these two chemicals are both present in the two first PCs. The loading profile of the `Time` effect shows a high peak on the left side of the removed water area.


#### Interaction `Hippurate:Time`

The scores and loadings fot the interaction term is represented here. It is not straighforward to interpret the scores plot of such an interaction term but the loadings of PC1 indicates a shift in the spectrum, along the whole spectral profile (but mostly around 3 ppm). 

```{r}
# Hippurate:Time
hiptim_scores_pl <- lmpScorePlot(resASCA, effectNames = "Hippurate:Time", 
                                 color = "Hippurate", shape = "Time")
hiptim_loadings_pl <- all_loadings_pl$`Hippurate:Time` + 
  time_peaks + 
  hip_peaks

grid.arrange(hiptim_scores_pl,hiptim_loadings_pl, ncol=2)
```

#### Combination of effects `Hippurate+Time+Hippurate:Time`

The scores and the loadings of a combination of effects, in this case `"Hippurate+Time+Hippurate:Time"` can also be rendered.

```{r}
# Hippurate+Time+Hippurate:Time
hiptimInter_scores_pl <- lmpScorePlot(resASCA, 
                            effectNames = "Hippurate+Time+Hippurate:Time", 
                            color = "Hippurate", shape = "Time")

hiptimInter_loadings_pl <- all_loadings_pl$`Hippurate:Time` + 
                            time_peaks + hip_peaks

grid.arrange(hiptimInter_scores_pl,hiptimInter_loadings_pl, ncol=2)
```

However, note that a better graphical representation is possible with the function `lmpEffectPlot()` (see Section \@ref(effects-plot)) to depict interaction terms or combinations of effects.


#### Model residuals

PCA on the model residuals can also be applied to spot outliers or an unexpected underlying structure of the data. The scores plot shows that the effect of `Day`, which was excluded in our modeling step, could have been included as well as it spans the two first PCs of the residuals. 

```{r Plot Residuals}

resid_scores_pl <- lmpScorePlot(resASCA, effectNames = "Residuals",
                                color = "Day", shape = "Day", 
                                drawShapes = "segment")


resid_loadings_pl <- all_loadings_pl$Residuals 

grid.arrange(resid_scores_pl,resid_loadings_pl, ncol=2)
```


### Other graphs

#### Scores scatter plot

We can also represent the scores as a matrix of plots with `lmpScoreScatterPlotM()`. This graph enables to observe multiple variables simultaneously. For example, the PC1 scores of `Hippurate`  and `Citrate` clearly represent the orthogonal design. the interaction `Hippurate:Time` can also be visulised when comparing PC1 of `Hippurate` and the interaction term.

```{r ASCA ScoresMatrix, fig.height=7}
lmpScoreScatterPlotM(resASCA,PCdim=c(1,1,1,1,1,1,1,2),
                     modelAbbrev = TRUE,
                     varname.colorup = "Citrate",
                     varname.colordown  = "Time",
                     varname.pchup="Hippurate",
                     varname.pchdown="Time",
                     title = "ASCA scores scatterplot matrix")
```

#### Loadings 2D

Finally the loadings could also be represented 2-by-2 as a scatterplot with `lmpLoading2dPlot()`. 

```{r ASCA loadings}
lmpLoading2dPlot(resLmpPcaEffects=resASCA, 
                 effectNames = c("Hippurate"), 
                 axes = c(1,2), 
                 addRownames = TRUE,pl_n = 10)

# adding manually labels to points for the Hippurate peaks
labels = substr(colnames(UCH$outcomes),1,4)
labels[-c(369, 132, 150, 133, 149, 144, 145, 368, 151)] <- ""

lmpLoading2dPlot(resLmpPcaEffects=resASCA, 
                 effectNames = c("Hippurate"), 
                 axes = c(1,2), points_labs = labels)
```


### Effects plot {#effects-plot}

The `lmpEffectPlot()` function is particularly interesting to visualise an interaction term or a combination of effects. Note that this function can only be applied with the `ASCA` method.


#### Main effects for `Hippurate`

Here the first PC of the `Hippurate` is represented along the levels of this effect.

```{r}
lmpEffectPlot(resASCA, effectName = "Hippurate", x = "Hippurate")
```


#### Interaction `Hippurate:Time`

A more interesting application is the visualisation of interaction terms as line plots, here `Hippurate:Time`. This representation gives the impression of a quite important interaction effect.

```{r}
lmpEffectPlot(resASCA, effectName = "Hippurate:Time", 
              x = "Hippurate", z = "Time")
lmpEffectPlot(resASCA, effectName = "Hippurate:Time", 
              x = "Time", z = "Hippurate")
```


#### Combination of effects `Hippurate+Time+Hippurate:Time`

An alternative visualisation combines the main effects of `Hippurate`, `Time` and the interaction `Hippurate:Time`. Even if significant, the interaction effect is actually quite small compared to the main effects (`Hippurate` for PC1 and `Time` for PC2).

```{r ASCA effects}
lmpEffectPlot(resASCA, effectName = "Hippurate+Time+Hippurate:Time", 
              x = "Hippurate", z = "Time")
lmpEffectPlot(resASCA, effectName = "Hippurate+Time+Hippurate:Time", 
              axes = c(2), x = "Time", z = "Hippurate")
```




## APCA

The APCA method performs PCA on the effect matrices augmented by the residuals. The same functions than for ASCA are used.

```{r APCA PCA}
resAPCA = lmpPcaEffects(resLmpEffectMatrices = resLmpEffectMatrices, 
                        method="APCA")
```

### Scores Plot

Different shapes with `drawShapes` argument highlight the data structure recovery.

```{r APCA ScoresXY}
# Hippurate main effect
lmpScorePlot(resAPCA, effectNames = "Hippurate", 
             color = "Hippurate", shape = "Hippurate", drawShapes = "ellipse")

# Citrate main effect
lmpScorePlot(resAPCA, effectNames = "Citrate", 
             color = "Citrate", shape = "Citrate", drawShapes = "ellipse")

# Time main effect
lmpScorePlot(resAPCA, effectNames = "Time", 
             color = "Time", shape = "Time", drawShapes = "ellipse")
lmpScorePlot(resAPCA, effectNames = "Time", 
             color = "Time", shape = "Time", drawShapes = "polygon")
lmpScorePlot(resAPCA, effectNames = "Time", 
             color = "Time", shape = "Time", drawShapes = "segment")

# Interaction term
lmpScorePlot(resAPCA, effectNames = "Hippurate:Time", 
             color = "Hippurate", shape = "Time", drawShapes = "segment")
lmpScorePlot(resAPCA, effectNames = "Hippurate:Time", 
             color = "Hippurate", shape = "Time", drawShapes = "polygon")
```

A scatterplot matrix can also be applied to visualise the relationship between the scores profiles of interest.

```{r APCA ScoresMatrix}
lmpScoreScatterPlotM(resAPCA,
                     effectNames = c("Hippurate", "Citrate", "Time",
                                     "Hippurate:Time"),
                     modelAbbrev = TRUE,
                     varname.colorup = "Citrate",
                     varname.colordown  = "Time",
                     varname.pchup="Hippurate",
                     varname.pchdown="Time",
                     title = "APCA scores scatterplot matrix")
```

### Loadings plot


Note that the APCA loadings contain the residuals of the model and differ from the ASCA loadings in that respect.

```{r APCA loadings}

lmpLoading1dPlot(resAPCA, effectNames = c("Hippurate", "Citrate", 
                                        "Time", "Hippurate:Time"), axes = 1)

```

## ASCA-E

The ASCA-E method performs PCA on the effect matrices then adds the residuals to compute the augmented scores. 

```{r ASCAE PCA}
resASCAE = lmpPcaEffects(resLmpEffectMatrices = resLmpEffectMatrices, 
                         method="ASCA-E")
```

Contributions and loadings are similar to the ASCA results.

### Scores Plot

For the main effects, all score plots show a clear separation between the different levels of the considered effects on the first PC. This separation of the groups suggests a strong effect of those factors, confirmed by their significance.

```{r ASCAE ScoresXY}
lmpScorePlot(resASCAE, effectNames = "Hippurate", 
             color = "Hippurate", shape = "Hippurate")
lmpScorePlot(resASCAE, effectNames = "Citrate", 
             color = "Citrate", shape = "Citrate")
lmpScorePlot(resASCAE, effectNames = "Time", 
             color = "Time", shape = "Time")
lmpScorePlot(resASCAE, effectNames = "Hippurate:Time", 
             color = "Hippurate", shape = "Time")
```

Again, the scores profiles can be compared 2 by 2 with ASCA-E.
```{r ASCAE ScoresMatrix}
lmpScoreScatterPlotM(resASCAE,
                     effectNames = c("Hippurate", "Citrate", "Time",
                                     "Hippurate:Time"),
                     modelAbbrev = TRUE,
                     varname.colorup = "Citrate",
                     varname.colordown  = "Time",
                     varname.pchup="Hippurate",
                     varname.pchdown="Time",
                     title = "ASCA-E scores scatterplot matrix")
```


# Session info
```{r}
sessionInfo()
```

# References
