---
title: "Application of the LMWiRe package on the UCH dataset"
author: "Benaiche Nadia, Sébastien Franceschini, Martin Manon, Thiel Michel, Govaerts Bernadette"
date: '`r format(Sys.time(), "%B %d, %Y,%H:%M")`'
package: LMWiRe
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    toc_float: true
vignette: >
   %\VignetteIndexEntry{Analysis of the UCH dataset with LMWiRe}
   %\VignetteEncoding{UTF-8}
   %\VignetteEngine{knitr::rmarkdown}
editor_options: 
   chunk_output_type: console
references:
- id: Thiel2017
  title: ASCA+ and APCA+ Extensions of ASCA and APCA in the analysis of unbalanced multifactorial designs
  author:
  - family: Thiel
    given: Michel
  - family: Feraud
    given: Baptiste
  - family: Govaerts
    given: Bernadette
  container-title: Journal of Chemometrics
  type: article-journal
  issued:
    year: 2017
- id: Guisset2019
  title: Comparison of PARAFASCA, AComDim, and AMOPLS approaches in the multivariate GLM modelling of multi-factorial designs
  author:
  - family: Guisset
    given: Severine
  - family: Martin
    given: Manon
  - family: Govaerts
    given: Bernadette
  container-title: Chemometrics and Intelligent Laboratory Systems
  type: article-journal
  issued:
    year: 2019
---

```{r Setup,include=FALSE}
require(knitr)
require(pander)
require(patchwork)
require(ggplot2)
knitr::opts_chunk$set(message=FALSE,warning=FALSE,comment=NA)
```

<span style="color: red;"> <font size="4"> THIS VIGNETTE IS STILL UNDER DEVELOPMENT (after section 6). IT WILL BE UPDATED SOON</font></span>

# Introduction

The purpose of this vignette is to show the possibilities offered by the LMWiRe package. 

LMWiRe stands for Linear Models for Wide Responses. This package was created to analyse models with wide responses and a multi-factor design of experiment and provide an implementation of ASCA and APCA derived methods.  

The model used in this example is a three-way ANOVA with fixed effects. This document presents all the usual steps of the analysis, from importing the data to visualising the results. Details on the methods used can be found in the articles of @Thiel2017 and @Guisset2019.

# Installation

The package is actually in its development stage and is available on GitHub at the url: <https://github.com/bgovaerts/LMWiRe/tree/develop>. It can be installed via the `remotes::install_github()` function. The package needs to be loaded afterwards.


```{r Install, results=FALSE, message=FALSE}
# if (!requireNamespace("remotes")) install.packages(pkgs="remotes", dependencies="Depends")
# remotes::install_github("ManonMartin/LMWiRe", dependencies = TRUE)
library("LMWiRe")
```

# Data importation

Before any analysis, the `UCH` data set needs to be loaded. The LMWiRe package contains the data set and it can be loaded with the `data()` function. The `load()` function is also useful to import your own data.

```{r Data Importation}

data("UCH")

```

# Data exploration

The UCH (Urine-Citrate-Hippurate) data set is described in @Thiel2017 and @Guisset2019 and is issued form a metabolomics experiment.  In this experiment, 36 samples of a pool of rat urine samples were spiked with two molecules Citrate and Hippurate according to a $3^2$ full factorial design in the quantities of these two molecules.  The spiked samples were analyzed by ^1^H NMR at two different time after defrozing and over two days.  Two of the spectra where finally missing at the end of the experiment.    

The UCH data set is a list containing 3 elements: 

* an `outcomes` matrix with 34 observations of 600 response variables representing the spectra from the \textsuperscript^1H-NMR$ spectroscopy, 
* a `design` matrix with 34 observations and 4 explanatory variables and 
* a `formula` for the General Linear Model (GLM) used. 

For the purpose of this example, only 3 factors of interest will be studied : Quantities of Hippurate and Citrate and Time after defrozing.

## Design

The design matrix contains the information about each observation for the four variables: Hippurate, Citrate, Day and Time. Only 3 of these variables are used in the model. The function `plotDesign` is useful to observe the design.

```{r Design visualization}

pander(head(UCH$design))
plotDesign(design = UCH$design, x = "Hippurate", 
           y = "Citrate", rows = "Time",
           title = "Design of the UCH dataset")

```

This plot confirms that the design is a full 3x3x2 factorial design replicated twice with 2 missing values.  The design is then not balanced. 

## Outcomes visualization

The 600 response (`outcomes`) variables represent, for each observation, the intensities of the ^1^H NMR spectra.   These spectra can be visualized by the `plotLine` function.

### `plotLine` function

Here, annotation are added to the `ggplot` in ordre to highligth the Hippurate (green) and Citrate (red) peaks. 

```{r Spectrum visualization}
cit_peaks <- annotate("rect", xmin=c(2.509), xmax=c(2.709), 
              ymin=-Inf, ymax=Inf, alpha=0.2, 
              fill=c("tomato")) 

hip_peaks <- annotate("rect", xmin=c(7.458,3.881), xmax=c(7.935,4.041), 
              ymin=-Inf, ymax=Inf, alpha=0.2, 
              fill=c("yellowgreen")) 

p1 <- plotLine(Y = UCH$outcomes,
         title = "H-NMR spectrum",
         rows = c(3),
         xlab = "ppm",
         ylab = "Intensity")

p1 + cit_peaks + hip_peaks
  
```

### `plotScatter` function

`plotScatter` function allows to visualize the values of two outcomes variables in highlighting with colors or markers the values of the design factors.  Here, it is used to show that the $3^2$ factorial design can be recovered from the intensities of the Hippurate and Citrate peaks in the spectra.  

```{r}

# xy corresponds to citrate (453) and hippurate peaks (369)
plotScatter(Y = UCH$outcomes,
            xy = c(453, 369),
            design = UCH$design,
            color = "Hippurate", 
            shape = "Citrate")
# Or
plotScatter(Y = UCH$outcomes,
            xy = c("2.6092056","3.9811536"),
            design = UCH$design,
            color = "Hippurate", 
            shape = "Citrate")

```

### `plotScatterM` function

`plotScatter` function allows to visualize the values of a series of outcomes variables in highlighting with colors or markers the values of the design factors.  It is done here for all peaks of Citrate et Hippurate

```{r}

plotScatterM(Y = UCH$outcomes, cols = c(133, 145, 150, 369, 453), 
             design = UCH$design,varname.colorup = "Hippurate", 
             varname.colordown = "Citrate")

```

### `plotMeans` function

`plotMeans` allows finally to visualize the mean values of a response variable for different levels of the design factors.  Here we show the evolution of the Citrate peak higth with respect to the three design factors of interest. 
Note that the results of this function must be interpreted with caution when designs are unbalanced.

```{r}

plotMeans(Y = UCH$outcomes,
          design = UCH$design,
          cols = c(453),
           x = c("Citrate"),
          w = c("Hippurate"),
          z = c("Time"),
          ylab = "Intensity",
          title=c("Mean reponse for main Citrate peak"))

```

## Principal Component Analysis

The function `pcaBySvd` is useful to compute a PCA decomposition of the `outcomes` matrix. The linked usefull functions are :

* `pcaScreePlot` to obtaine a scree plot
* `pcaLoadingPlot` for the loading plots
* `pcaScorePlot` for the score plots

```{r PCA}

ResPCA = pcaBySvd(UCH$outcomes)
pcaScreePlot(ResPCA, nPC = 6) 

```

The score plots show a quite effect of the three design factors which will be more clearly highlighted by ASCA and APCA.  

```{r Scores}

pcaScorePlot(resPcaBySvd = ResPCA, axes = c(1,2), 
             title = "PCA scores plot: PC1 and PC2", 
             design = UCH$design,
             color = "Hippurate", shape = "Citrate",
             points_labs_rn = FALSE)

pcaScorePlot(resPcaBySvd = ResPCA, axes = c(1,2), 
             title = "PCA scores plot: PC1 and PC2", 
             design = UCH$design,
             color = "Time", shape = "Hippurate",
             points_labs_rn = FALSE)

pcaScorePlot(resPcaBySvd = ResPCA, axes = c(3,4), 
             title = "PCA scores plot: PC3 and PC4", 
             design = UCH$design,
             color = "Time", shape = "Citrate",
             points_labs_rn = FALSE)

```

In the first two loading plots a mix of Citrate and Hippurate peaks already appear.  

```{r Loadings}

p2 <- pcaLoadingPlot(resPcaBySvd = ResPCA, axes = c(1,2), 
               title = "PCA loadings plot UCH", xlab = "ppm", 
               ylab = "Intensity")

p2 +  hip_peaks + cit_peaks

```

# Model estimation and effect matrix decomposition

## Model formula

The formula of the ANOVA-GLM model used in this analysis is the 3 ways crossed ANOVA model:

```{r Formula}
UCH$formula
```

## Model matrix generation

The first step of ASCA+ is to build the (GLM) model matrix from the experimental design matrix and the model. Each factor is reencoded with multiple binary variables using  `contr.sum` coding. The model matrix is a \emph(34xp) with p being the total number of parameter in the ANOVA model for one response.

The function `lmwModelMatrix()` encodes the design matrix as a model matrix.

```{r ModelMatrix}

resLmwModelMatrix = lmwModelMatrix(UCH)
pander::pander(head(resLmwModelMatrix$modelMatrix))
```

## Model estimation and effect matrices decomposition

`lmwEffectMatrices()` is the used to estimate the GLM model and decompose the outcomes matrix effect matrices for every model term.  This function calculates also type III effect contributions (in %) and generates a linked `barpot`.  

```{r EffectMatrices}
resLmwEffectMatrices = lmwEffectMatrices(resLmwModelMatrix)
```

### Print and plot effect contributions

```{r}
pander(resLmwEffectMatrices$variationPercentages)
resLmwEffectMatrices$varPercentagesPlot
```

## Bootstrap tests and quantification of effects importance

`lmwBootstrapTests()` allows to apply parametric bootstrap test to determine whether an effect is significant or not.  Use first a small value of `nboot` (e.g. nboot=100) to develop your code and increase then it (e.g. nboot=1000) in ordre to get an accurate value for the p-values.    

```{r Bootstrap}

resLmwBootstrapTests = 
  lmwBootstrapTests(resLmwEffectMatrices = resLmwEffectMatrices, 
                                         nboot=100)

# Print P-values
pander::pander(t(resLmwBootstrapTests$resultsTable))

```

# ASCA/APCA/ASCA-E decomposition

ASCA/APCA/ASCA-E decomposition allow to represent the information from the effect matrices in a space of reduced dimensions. The function `lmwPcaEffects()` has a method argument to define which method to use.

## ASCA

The ASCA method performs PCAs on the pure effect matrices.

```{r ASCA PCA}
resASCA = lmwPcaEffects(resLmwEffectMatrices = resLmwEffectMatrices, 
                        method="ASCA", 
                        combineEffects = list(c("Hippurate", "Time", 
                                                "Hippurate:Time")))
```

### Contributions

The contributions from each effect and their components are estimated and reported in tables with the function `lmwContributions()`. Moreover the function also output barplots with the ordered contributions.

```{r ASCA Contrib}
resLmwContributions = lmwContributions(resASCA)

pander::pander(resLmwContributions$totalContribTable)
pander::pander(resLmwContributions$effectTable)
pander::pander(resLmwContributions$contribTable)
pander::pander(resLmwContributions$combinedEffectTable)

resLmwContributions$plotContrib + ggplot2::geom_bar(stat = "identity", 
                                                    fill = "cadetblue4")
```

### Scores Plot

The score matrices are then represented on two components with the function `lmwScorePlot()`.

```{r ASCA ScoresXY}
lmwScorePlot(resASCA, effectNames = "Hippurate", 
             color = "Hippurate", shape = "Hippurate")
lmwScorePlot(resASCA, effectNames = "Citrate", 
             color = "Citrate", shape = "Citrate")
lmwScorePlot(resASCA, effectNames = "Time", color = "Time", shape = "Time")
lmwScorePlot(resASCA, effectNames = "Hippurate:Time", 
             color = "Hippurate", shape = "Time")
lmwScorePlot(resASCA, effectNames = "Hippurate+Time+Hippurate:Time", 
             color = "Hippurate", shape = "Time")
lmwScorePlot(resASCA, effectNames = "Hippurate+Time+Hippurate:Time", 
             color = "Hippurate", shape = "Time", axes = c(3,4))
```

Finally we can represent the scores with a matrix of plots. This graph allows to observe multiple variables simultaneously.

```{r ASCA ScoresMatrix, fig.height=8}
lmwScoreScatterPlotM(resASCA,PCdim=c(1,1,1,1,1,1,1,2),
                     modelAbbrev = TRUE,
                     varname.colorup = "Citrate",
                     varname.colordown  = "Time",
                     varname.pchup="Hippurate",
                     varname.pchdown="Time",
                     title = "ASCA scores scatterplot matrix")
```

### Effects plot

```{r ASCA effects}
lmwEffectPlot(resASCA, effectName = "Hippurate", x = "Hippurate")
lmwEffectPlot(resASCA, effectName = "Citrate", x = "Citrate")
lmwEffectPlot(resASCA, effectName = "Time", x = "Time")
lmwEffectPlot(resASCA, effectName = "Hippurate:Time", 
              x = "Hippurate", z = "Time")
lmwEffectPlot(resASCA, effectName = "Hippurate:Time", 
              x = "Time", z = "Hippurate")
lmwEffectPlot(resASCA, effectName = "Hippurate+Time+Hippurate:Time", 
              x = "Hippurate", z = "Time")
lmwEffectPlot(resASCA, effectName = "Hippurate+Time+Hippurate:Time", 
              axes = c(1:3), x = "Time", z = "Hippurate")
```


### Loadings plot

```{r ASCA loadings}
time_peaks <- annotate("rect", xmin=c(5.955364), xmax=c(6.155364), 
              ymin=-Inf, ymax=Inf, alpha=0.2, 
              fill=c("royalblue")) 

resLoadASCA <- lmwLoadingPlot(resASCA, 
                              effectNames = c("Hippurate", "Citrate","Time",
                                              "Hippurate:Time",
                                              "Hippurate+Time+Hippurate:Time"), 
                              axes = 1, xlab = "ppm")
resLoadASCA[[1]] +  hip_peaks

resLoadASCA[[2]] + cit_peaks

resLoadASCA[[3]] + time_peaks

resLoadASCA[[4]] + time_peaks + hip_peaks

resLoadASCA[[5]] + time_peaks + hip_peaks

```

### Residuals plot

```{r Plot Residuals}

lmwScorePlot(resASCA, effectNames = "Residuals", color = "Day", 
             shape = "Day", drawShapes = "ellipse")

```


## APCA

The APCA method performs PCAs on the effect matrices augmented by the residuals. The same functions are used.

```{r APCA PCA}
resAPCA = lmwPcaEffects(resLmwEffectMatrices = resLmwEffectMatrices, 
                        method="APCA")
```

### Contributions

```{r APCA Contrib}
resLmwContributions = lmwContributions(resAPCA)

pander::pander(resLmwContributions$effectTable)
```

### Scores Plot

```{r APCA ScoresXY}
lmwScorePlot(resAPCA, effectNames = "Hippurate", 
             color = "Hippurate", shape = "Hippurate", drawShapes = "ellipse")
lmwScorePlot(resAPCA, effectNames = "Citrate", 
             color = "Citrate", shape = "Citrate", drawShapes = "ellipse")

lmwScorePlot(resAPCA, effectNames = "Time", 
             color = "Time", shape = "Time", drawShapes = "ellipse")
lmwScorePlot(resAPCA, effectNames = "Time", 
             color = "Time", shape = "Time", drawShapes = "polygon")
lmwScorePlot(resAPCA, effectNames = "Time", 
             color = "Time", shape = "Time", drawShapes = "segment")

lmwScorePlot(resAPCA, effectNames = "Hippurate:Time", 
             color = "Hippurate", shape = "Time", drawShapes = "segment")
lmwScorePlot(resAPCA, effectNames = "Hippurate:Time", 
             color = "Hippurate", shape = "Time", drawShapes = "polygon")
```

```{r APCA ScoresMatrix}
lmwScoreScatterPlotM(resAPCA,
                     effectNames = c("Hippurate", "Citrate", "Time",
                                     "Hippurate:Time"),
                     modelAbbrev = TRUE,
                     varname.colorup = "Citrate",
                     varname.colordown  = "Time",
                     varname.pchup="Hippurate",
                     varname.pchdown="Time",
                     title = "APCA scores scatterplot matrix")
```

### Loadings plot

```{r APCA loadings}

lmwLoadingPlot(resAPCA, effectNames = c("Hippurate", "Citrate", 
                                        "Time", "Hippurate:Time"), axes = 1)

```

## ASCA-E

The ASCA-E method performs PCAs on the effect matrices then add the residuals. The same functions are used.

```{r ASCAE PCA}
resASCAE = lmwPcaEffects(resLmwEffectMatrices = resLmwEffectMatrices, 
                         method="ASCA-E")
```

Contributions and loadings are identical to the ASCA results.

### Scores Plot

```{r ASCAE ScoresXY}
lmwScorePlot(resASCAE, effectNames = "Hippurate", 
             color = "Hippurate", shape = "Hippurate")
lmwScorePlot(resASCAE, effectNames = "Citrate", 
             color = "Citrate", shape = "Citrate")
lmwScorePlot(resASCAE, effectNames = "Time", 
             color = "Time", shape = "Time")
lmwScorePlot(resASCAE, effectNames = "Hippurate:Time", 
             color = "Hippurate", shape = "Time")
```

```{r ASCAE ScoresMatrix}
lmwScoreScatterPlotM(resASCAE,
                     modelAbbrev = TRUE,
                     effectNames = c("Hippurate", "Citrate", "Time",
                                     "Hippurate:Time"),
                     varname.colorup = "Citrate",
                     vec.colorup = c("royalblue","yellowgreen","lightcoral"),
                     varname.colordown  = "Time",
                     vec.colordown = c("cadetblue4","deeppink4"),
                     varname.pchup="Hippurate",
                     vec.pchup = c(4,16,2),
                     varname.pchdown="Time",
                     vec.pchdown = c(1,3), 
                     title = "ASCA-E scores scatterplot matrix")
```


# Session info
```{r}
sessionInfo()
```

# References
